Non-parametric approaches to optimal policy are more robust

### 2013-03-21 inst/doc/beyond-mechanistic-models.Rmd
pomp example

```{r}
read_chunk("nonparametric-bayes.R")
```
```{r knit-settings, echo=FALSE, include=FALSE, cache=FALSE}
```
```{r libraries, echo=FALSE, include=FALSE}
```
```{r plotting-options, echo=FALSE, include=FALSE}
```
```{r stateeq}
```
```{r sdp-pars, dependson="stateeq"}
```
```{r obs, dependson="sdp-pars"}
```
```{r mle, dependson="obs"}
```
```{r gp-priors}
```
```{r gp, dependson=c("obs", "gp-priors")}
```
```{r opt, dependson=c("gp", "mle")}
```
```{r sim, dependson="opt"}
```

@@

@@

Mechanistic modeling emphasizes the importance of capturing the correct
gross properties of a system over the tracking minute fluctuations.
For instance, in selecting and parameterizing model of a population
of conservation concern, we may be most interested in getting the
long-term behavior correct -- such as identifying if the dynamics support
persistence of the population -- rather than worrying how well they
reflect the year-to-year fluctuations.  We would certainly have good
reason to prefer such a model over alternatives which are irreconcilable
to the most basic biological processes, such as unbounded growth, or
growth curves that do not pass through the origin in a closed system. So
it may come as a surprise to realize that such obviously wrong models
can perform as well or even better than reasonable mechanistic models
in guiding ecological management and decision making.

@@

@@

Three elements contribute to this.

A process-based understanding helps us formulate hypotheses, guide
experimental design and data collection, predict outcomes, and influence
management decisions.  The alternative to a processed-based model is  a
correlative one. Statical models machine learning.

@@

machine learning algorithms

Prediction without

Example of Statistical model vs process-based model.

>>>>

% Beyond Mechanistic Models: The case for machine learning in ecological management

Abstract
============

<!-- Run R code for analysis, to be called by figures -->
```{r, echo=FALSE, cache=FALSE, include=FALSE}
options(xtable.print.comment=FALSE)
options(xtable.type = 'latex', table.placement="H")
opts_chunk$set(warning=FALSE, message=FALSE, comment=NA,
               tidy=FALSE, echo=FALSE)
opts_knit$set(progress = TRUE, verbose = TRUE)
opts_chunk$set(dev = 'Cairo_pdf', fig.width=5.5, fig.height=4,
 cache.path = 'cache-pdf/', cache=TRUE, external=TRUE)
read_chunk("nonparametric-bayes.R")
```
```{r knit-settings, echo=FALSE, include=FALSE, cache=FALSE}
```
```{r libraries, echo=FALSE, include=FALSE}
```
```{r plotting-options, echo=FALSE, include=FALSE}
```
```{r stateeq}
```
```{r sdp-pars, dependson="stateeq"}
```
```{r obs, dependson="sdp-pars"}
```
```{r mle, dependson="obs"}
```
```{r gp-priors}
```
```{r gp, dependson=c("obs", "gp-priors")}
```
```{r opt, dependson=c("gp", "mle")}
```
```{r sim, dependson="opt"}
```

Introduction
===============================================================

@@

Mechanistic modeling emphasizes the importance of capturing the correct
gross properties of a system over tracking minute fluctuations.
For instance, in selecting and parameterizing model of a population
of conservation concern, we may be most interested in getting the
long-term behavior correct -- such as identifying if the dynamics support
persistence of the population -- rather than worrying how well they
reflect the year-to-year fluctuations.  We would certainly have good
reason to prefer such a model over alternatives which are irreconcilable
to the most basic biological processes, such as unbounded growth, or
growth curves that do not pass through the origin in a closed system. So
it may come as a surprise to realize that such obviously wrong models
can perform as well or even better than reasonable mechanistic models
in guiding ecological management and decision making.

@@

Models & Methods
===========================================================================

<!-- Include example of a machine learning prediction that involves biologically
inaccurate predictions such as unbounded growth or growth at zero and yet still
performs well (by fitting well in the region of interest?)  -->

@@

Discussion
===========================================================================

@@

@@

### 2013-03-21 inst/doc/nonparametric-bayes.Rmd
some writing
pomp example

* Most management recommendations from the ecological literature
/ management policies based on (motivated by) parametric models.
_Preference to pitch towards policy or theoretical literature?_

* Background discussion on the importance/success of parametric modeling;
(e.g. @Levins1966, etc., up through @Geritz2011a)

* Background on the concerns of structural uncertainty -- we don't have
the right models.  (Also: measurement uncertainty, parameter uncertainty,
unobserved states, boundary conditions, etc.)

* Hierarchical Bayesian approach has provided a natural way to address
these from a statistical standpoint.  Successes and challenges from the
parametric route: e.g. @Cressie2009.

* Among pitfalls of these approaches: particularly difficult to apply in
management context. Optimization-based (Decision-theoretic) approaches to
ecological management need to be precisely parameterized about everything,
even the uncertainty itself [@Polasky2011] (discussion of @Brozovic2011
SDP example in context of threshold system would perhaps be useful
as well).

@@

* Nonparametric approaches are beginning to appear more frequently
in ecological and conservation literature (Species distribution
models/maxent), including the Gaussian process based approach used here
[@Munch2005].  However, such approaches have yet to be applied to the
decision-theoretic framework that could guide management decisions,
where we expect them to excel for several reasons (1) the ability to
make accurate forecasts by more closely approximating the underlying
process where data is available (2) remaining appropriately ambiguous
where data is not available (3) remaining computationally simple enough
to avoid some pitfalls common to hierarchical parametric approaches.

@@

The use of Gaussian process regression (known as Kreging in the geospatial literature) to formulate a predictive model is relatively new in the context of modeling dynamical systems [@Kocijan2005] and introduced in the ecological modeling and fisheries management by @Munch2005.

_Is the following  discription useful or too elementary?_

The essense of the Gaussian process approach can be captured in the following thought experiment: An exhaustive parametric approach to the challenge of structural uncertainty might proceed by writing down all possible functional forms for the underlying dynamical system with all possible parameter values for each form, and then consider searching over this huge space to select the most likely model and parameters; or using a Bayesian approach, assign priors to each of these possible models and infer the posterior distribution of possible models. The Gaussian process approach can be thought of as a computationally efficient approximation to this approach. Gaussian processes represent a large class of models that can be though of as capturing or reasonably approximating the set of models in this collection.  By modeling at the level of the process, rather than the level of parametric equation, we can more concisely capture the possible behavior of these curves.  In place of a parametric model of the dynamical system, the Gaussian Process approach postulates a prior distribution of (n-dimensional) curves that can be though of as approximations to a large range of possible

_Do we need more specifics on Gaussian process as an approximation to parametric models? Discussion of Gaussian process vs other machine learning / forecasting approaches that have less clear statistical foundations?  If so, does this all belong in the discussion?_

The Gaussian process is defined by a covariance kernel ... _discussion of our (gaussian) kernel, blah blah how specific here vs appendix?_.  By requiring our kernel to follow a generic functional form, we can compactly describe the Gaussian process using only a handful of parameters (Table 1)

A more thorough introduction to the formulation and use of Gaussian processes can be found in @Rasmussen2006.

* Gaussian processes in ecological literature [@Munch2005]

* Our methodology (e.g. following @Munch2005).

###  Discussion of the dynamic programming solution

@@

>>>>

Most management recommendations from the ecological literature are based on
(or at least motivated by) parametric models. Though in principle these models
reflect mechanistic underpinnings of biological interactions involved, in practice
real mechanisms are often unknown and the true dynamics too complex to be captured
by simple models [@refs]. While simple mechanistic models can nevertheless
provide imporant insights into possible dynamics -- for instance, demonstrating
that a vaccine does not have to be 100% effective to eliminate the transmission of a virus [@Kermack1921] -- such approaches are not well-suited for use in 
forecasting future dynamics upon which outcomes management policy can be based.  
Despite a long history of work emphasizing the difference between modeling 
for understanding (generic or strategic models) and modeling for prediction (precise, specific, or tactical models) [e.g. @Levins1966], simple parametric models 
continue to find application in predictive management contexts.  The management
of marine fisheries is a prime example, in which simple, mechanistically 
motivated models of stock-recruitment relationships such as Ricker or Beverton-Holt
curves are systematically fit to data and used in solving decision-theoretic
or optimal control problems determining resource management policies such
as maximum sustainable yield [@refs].

Stochastic dynamic programming @Mangel1982

To address structural uncertainty in the processes involved, such mechanistic models
can be adapted to accomidate uncertainty in the parameters or consider a set of 
alternative models simultaneously (belief SDP)

Background on the concerns of structural uncertainty -- we don't have
the right models.  Further unknonws such as measurement uncertainty,
parameter uncertainty, unobserved states, knowledge of boundary
conditions, etc. further compound the issue.  Though a hierarchical
Bayesian approach provides a natural way to address these from a
statistical standpoint, formulating reasonable parametric descriptions
of each form of uncertainty is a challenging task in itself, let alone
the computational difficulties of solving such a system. @Cressie2009
provides a good account of the successess and challenges of the approach. 
Applying these approaches in the management context of sequential decision
making, in which forecasts must be obtained over a range of possible actions
and updated regularly as new information arrives makes such an approach 
less feasible still.

@@

Though machine learning approaches have begun to appear in the
ecological and conservation literature (Species distribution models),
including the Gaussian process based approach used here [@Munch2005],
they remain unfamiliar and untrusted approaches for most ecologists.
Machine learning approaches represent an essentially pattern-based
rather than process-based approach, raising the same skepticism from most
theoretical ecologists that they hold for older correlative methods such
as linear regression, while the complexity of the techniques has barred
their adoption in more empirical audiences. <!-- Whoa! rampant speculation
alert! -->

.  However, such approaches have yet to be applied to the
decision-theoretic framework that could guide management decisions,
where we expect them to excel for several reasons (1) the ability to
make accurate forecasts by more closely approximating the underlying
process where data is available (2) remaining appropriately ambiguous
where data is not available (3) remaining computationally simple enough
to avoid some pitfalls common to hierarchical parametric approaches.

@@

The use of Gaussian process regression (known as Kreging in the geospatial literature) to formulate a predictive model is relatively new in the context of modeling dynamical systems [@Kocijan2005] and introduced in the ecological modeling and fisheries management by @Munch2005.  An accessible and thorough introduction to the formulation and use of Gaussian processes can be found in @Rasmussen2006.

The essense of the Gaussian process approach can be captured in the following thought experiment: An exhaustive parametric approach to the challenge of structural uncertainty might proceed by writing down all possible functional forms for the underlying dynamical system with all possible parameter values for each form, and then consider searching over this huge space to select the most likely model and parameters; or using a Bayesian approach, assign priors to each of these possible models and infer the posterior distribution of possible models. The Gaussian process approach can be thought of as a computationally efficient approximation to this approach. Gaussian processes represent a large class of models that can be though of as capturing or reasonably approximating the set of models in this collection.  By modeling at the level of the process, rather than the level of parametric equation, we can more concisely capture the possible behavior of these curves.  In place of a parametric model of the dynamical system, the Gaussian Process approach postulates a prior distribution of (n-dimensional) curves that can be though of as approximations to a range of possible (parametric) models that might describe the data. The GP allows us to consider a set of possible curves simultaneously (Figure 1).  <!-- Figure of curves drawn from the posterior density? -->

<!-- Do we need more specifics on Gaussian process as an approximation to parametric models? Discussion of Gaussian process vs other machine learning / forecasting approaches that have less clear statistical foundations?  If so, does this all belong in the discussion? -->

### The optimal control problem

@@

### The Guassian process

The Gaussian process is defined by a covariance kernel.  By requiring our kernel to follow a generic functional form, we can compactly describe the Gaussian process using only a handful of parameters (Table 1)

###  Formulating a dynamic programming solution
